Ideas

### 2023-07-17 Notes.tex
typo

In particular, if one just take the definition with general
interchange pairs, undesidered things might happen. In particular one
can have shift-equivalent derivations where the permutation is the
identity but which are non isomorphic (e.g., in GraphShift, page 1-2
$\rho_1 \shifteq \rho_3$, but $\rho_1 \equiv^{a} \rho_3$.

We want a notion of shift-equivalence that ensures some properties
which are indispensable to allow to interpret derivations up-to shift
as equivalent concurrent computations which differ only for the fact
that independent steps have been moved around. When moving around steps, we should not lose track of the identity of the steps: this is essential to be able to have an event structure semantics with a clearly defined notion of event.

Hence we assume to deal
with a notion of shift-equivalence between derivations of the kind
$\rho \shifteq_\sigma \rho'$ where $\rho, \rho'$ consist of $n$
rewriting steps, $\sigma: \interval{n} \to \interval{n}$ is a
permutation (the idea being that $\rho$ is obtained from $\rho'$
pemutating the steps according to $\sigma$) with the following
properties:

\begin{itemize}
\item \textbf{Property 0}. Shift equivalence is an equivalence, in particular it is symmetric. I.e., if  $\rho \shifteq_\sigma \rho'$ then $\rho' \shifteq_{\sigma^{-1}} \rho$;

@@

\textbf{Question.} Taking the standard shift with generic interchange pair, a counterexample to Property 1 is the one mentioned above.  Do we have a counterexample to Property 2? Not clear to me.

@@

\item For left-linear rules, the proposed notion of shift equivalence
  requires: uniqueness of the interchange pair and symmetry. Show that
  this satisfies the above

\item Apparently, the satisfaction of the properties above has not
  been explicitly observed for the case of linear rules. If so, it is
  worth to remark it.

\item Does the theory above apply to other cases. E.g., if one wants a
  proper notion of causality (avoiding or-causality) a possibility
  seems to be to restrict the applicability condition for rules, to
  prevent fusions to act on already merged items (see section V.D in
  UnstableFull). By the way, note that the satisfaction of the conditions
  on shift depends not only on the notion of shift, but also on the
  notion of rewriting. In this case we expect to have a prime event
  structure (but this has to do with the second part of the work).

@@

Ideally, one would like to be able to work with an abstract notion of
shift-equivalence and just using its properties as singled out in the
previous part to derive a weak prime domain or connected event structure
for it. This is probably difficult, but one might need to add some
futher properties to be proved for the specific notion of shift at
hand.

For this part, relevant material is section V of UnstableFull. We use abstraction equivalence (iso of derivations) $\equiv^{a}$ (Def. 36), shift-equivalence $\shifteq_\sigma$ (Def. 38, traking the permutation) and their join $\equiv^c_\sigma$.

For linear rules on graphs it is known that two derivations are
related by $\rho \equiv^c_\sigma \rho'$ iff $\sigma$ is a consistent
permutation (Def. 40, UnstableFull). Roughly the idea is to take the colimit
and consider the rewriting steps over the colimit to deduce
dependencies. This was used in many places in the proofs, but for
left-linear rules the proof does not go through (only the implication
``if $\rho \equiv^c_\sigma \rho'$ then $\sigma$ is a consistent
permutation'' (Lemma 17(1)) seems to hold as the original proof do not
to rely on linearity, this should be double-checked, but the converse
is false). In general, the hope is to get rid of using consistent
permutations and use instead more abstract properties of
shift-equivalence as discussed in the first part.

The general idea is to show that considering derivations are arrows from the source to the target graph, then the slice from the initial graph $\slice{[G_s]}{\tr{\mathcal{G}}}$ should be a pre-order, whose ideal completion is a weak prime domain.

@@

\keywords{Shift equivalence, non-linear rules, DPO rewriting.}

\section{Introduction}
\label{se:graphs}

Shortly, shift equivalence can be defined as the smallest equivalence relation on derivations, i.e. sequences of rewriting steps,
that can be obtained by the iterated switch of sequentially 
independent steps. A crucial observation, which represented our starting point, is that the notion of sequential independence traditionally
adopted for linear graph transfomation systems is not adequate when we
allow rules to be non-right linear, i.e., in the presence of
fusions. Consider the rules in Fig.~\ref{fi:rules}, typed over the type graph $T$,
where numbers are used to represent the mappings from the interface to the left- and
right-hand sides.

\begin{figure}
  \input{Figure/rules.tex}

\caption{The type graph  and the rules of the grammar of the example.}
\label{fi:rules}
\end{figure}

Now, let us consider an initial graph with a single node. We can then obtain
the derivation $\rho_1 = \delta_1; \delta_2; \delta_3$ in Fig.~\ref{fi:der1}, where $\delta_2$ and $\delta_3$ are
sequential independent according to the standard notion.

\begin{figure}
  \input{Figure/der1.tex}
\caption{The derivation $\rho_1$.}
\label{fi:der1}
\end{figure}

We can exchange the applications of $p_2$ and $p_3$, thus getting the
derivation $\rho_2 = \delta_1; \delta_3'; \delta_2'$ in
Figure~\ref{fi:der2}. Now, one can exchange again $p_3$ and $p_2$,
reaching the derivation $\rho_3 = \delta_1; \delta_2'; \delta_3$ in
Figure~\ref{fi:der3}, where in $\delta_2'$ production $p_2$ uses the
node of the start graph and thus ``does not depend'' on $p_1$.

\begin{figure}[t]
  \input{Figure/der2.tex}
\caption{The derivation $\rho_2$.}
\label{fi:der2}
\end{figure}

\begin{figure}[t]
  \input{Figure/der3.tex}
\caption{The derivation $\rho_3$.}
\label{fi:der3}
\end{figure}

The fact that if we switch two steps and then switch them back, we obtain a derivation that is not equivalent to the original one has unpleasant consequences. For instance, if we restrict $\rho_1'$ and $\rho_3'$ to the first two steps, we get non-equivalent derivations $\delta_1;\delta_2 \not\shifteq \delta_1;\delta_2'$, but $\rho_1 \shifteq[\sigma] \rho_3$, with $\sigma$ mapping $\delta_1$ to $\delta_1$ and $\delta_2$ to $\delta_2'$.
%
The lack of such closure with respect to composition does not allow to obtain a causal semantics for graph transformation systems.
More precisely, it does  not allow to associate an event structure to a derivation.
%
Thus, the notion of sequential independence has to be refined. The point is that in the associated shift equivalence we want to forget who made the fusion, 
but we should not forget (as it happens in the example) that some fusions occurred.

% In Andrea's counterexample (same as above, but $p_2$ produces an item that is consumed by $p_3$, thus forcing $p_3$ to happen last), if we construct the domain we have:
% \begin{center}
%   \includegraphics[scale=.7]{dom.pdf}
% \end{center}
% which is a weak prime domain (actually a prime domain) but it does not capture the proper semantics (e.g., $p_2'$ and $p_3$ are the same event!).

Shift equivalence is made stricter by asking that two steps are
sequentially independent if the result of their switch is uniquely
determined. Intuitively, when there are several way of performing the
exchange, it means that the first step has executed some fusions that
the second step is using in an essential way, hence they should not
considered independent. Formally, this boils down to require that 
in the definition of sequential independence the interchange pair is unique.

The paper builds on this intuition: it introduces this stricter notion of sequential independence
(Section~xxx) and prove that the associated theory of shift equivalence on derivations is 
well-behaved with respect to the concatenation of rewriting steps (Section~xxx).
Finally, it outlines how the relationship with event structure works, wrapping up the paper
with comparison with related woks and hinting at further direction of research (Section~xxx).

\section{Preliminaries [to be redone]}

\begin{figure}[t]
   \input{Figure/basic_independence}
\caption{Sequential independence for
  $\rho = G \Rrel{p_1/m_1} H \Rrel{p_2/m_2} M$.}
\label{fi:strongseq}
\end{figure}

\begin{definition}[sequential independence]
  \label{de:seq-ind}
  Consider a derivation $G \Rrel{p_1/m_1} H \Rrel{p_2/m_2} M$ as in
  Fig.~\ref{fi:strongseq}. Then, its components are \emph{sequentially
    independent} if there exists a unique \emph{independence pair} among
  them, i.e., two graph morphisms $i_1: R_1 \rightarrow D_2$ and
  $i_2: L_2 \rightarrow D_1$ such that $l_2^* \circ i_1 = m_{R_1}$,
  $r_1^* \circ i_2 = m_{L_2}$.
\end{definition}

Observe that this notion reduces to the usual one for linear rules. In
fact when $r_1^*$ and $l_2^*$ are mono the uniqueness requirement is
trivially satisfied. Instead, for left-injective rules $i_1$ is still
unique, but $i_2$ might be not (as it happens, e.g., in the example
before).\todo{Say more about the example}

Observe that the notion is not symmetric: if
$\rho = G \Rrel{p_1/m_1} H \Rrel{p_2/m_2} M$ is sequential independent
and we switch it, we could obtain a pair which is not because it fails
to satisfy uniqueness of the independence pair (this happens, e.g., in
our example, where $\delta_2;\delta_3$ is sequential independent, but
$\delta_3';\delta_2'$ is not). The idea is to define shift equivalence
as $\rho \equiv_{sh} \rho'$ when \emph{each of the two} can be
obtained from the other by exchanging sequential independent
derivations. More precisely, we have a shift relation, which is used
for defining the equivalence as in Fig.~\ref{fi:ind}. It could be
better to have a slightly less elegant but more direct
definition. Write $\rho \shiftdir[\transp{i}] \rho'$ when
$\rho = \rho_1; \rho_2; \rho_3$ and $\rho' = \rho_1; \rho_2', \rho_3$
with $|\rho_1| = i-1$, $|\rho_2|=2$ and $\rho_2' \in
\IC{\rho_2}$. Then we write $\rho \shiftpre[\perm] \rho'$ if
$\rho = \rho_0 \shiftdir[\perm_1] \rho_1 \ldots \shiftdir[\perm_n ]
\rho'$ with $\perm = \perm_1; \ldots; \perm_n$ for some $n \geq 0$
(with the convention that for $n=0$ we get the identity).

The restricted notion of sequential independence ensures that the interchange operator produces a result that is unique up to abstraction equivalence.

\begin{lemma}[IC is unique]
  \label{le:unique-IC} 
  Let $\rho$ be a derivation such that $|\rho|=2$. If $\rho$ is
  sequential independent, let us write
  $\IC{\rho} = \{ \rho' \mid \rho \shiftdir \rho'\}$. Then for
  all $\rho', \rho'' \in \IC{\rho}$ we have $\rho' \equiv
  \rho''$. Moreover, if $\rho$ is sequential independent then for all
  $\rho' \in \IC{\rho}$, if $\rho'$ is sequential independent and
  $\rho'' \in \IC{\rho'}$, we have $\rho \equiv \rho''$.
\end{lemma}

\begin{figure}[t]
  \input{Figure/independence}
  \caption{Independence.}
  \label{fi:independence}
\end{figure}

It is convenient (at the beginning, at least) to work with shift
on derivations where the source and target graph is concrete and the
middle can be taken up to iso. I think that everything can be then
easily transferred to ctc-equivalence. We say that $\rho$ and $\rho'$
are abstract shift-equivalent and write $\rho \sim \rho'$.

A crucial observation is that independence can be defined also for non-contiguous steps.

\begin{definition}[independence]
  Let $\delta_1;\rho;\delta_n$ be a derivation as in
  Fig.~\ref{fi:independence}, where $\delta_1$ and $\delta_n$ are
  direct derivations and $\rho$ is a generic, possibly empty
  derivation. We say that $\delta_1$ and $\delta_n$ are independent if
  there is a unique independence pair $i_{n,1}$ and $i_{1,n}$
  satisfying the expected commutation, where $C$ is the colimit of the
  bottom line (excluding $G_1, G_{n+1}$).
\end{definition}

The fact that the notion of independence makes sense is proved by the following lemma, showing that the notion is preserved by the shift relation.

\begin{lemma}[independence is global]
  \label{le:seq-ind-glob1}
  The notion of independence for contiguous derivation steps
  coincides with sequential independence.

  Moreover, independence is invariant under the shift-relation, i.e.,
  \begin{itemize}

  \item if $\delta_1;\rho;\delta_2$ and $\rho \shiftpre \delta'$ then
    in $\delta_1;\rho;\delta_2$ the steps $\delta_1$, $\delta_2$ are
    independent iff they are in $\delta_1;\delta';\delta_2$.

  \item if $\delta_1;\rho;\delta';\delta_2$ and
    $\delta_2'; \delta'' \in \IC{\delta';\delta_2}$ then in
    $\delta_1;\rho;\delta';\delta_2$ the steps $\delta_1$ and $\delta_2$ are
    independent iff they are in $\delta_1;\rho;\delta_2'$.

  \item same as before, but with shift on the left.
  \end{itemize}
\end{lemma}

\begin{proof}
  To be done. The fact that sequential independence coincides with
  independence for contiguous derivations is easy. The fact that it coincides when derivations are not contiguous is false (the approach based on defining independence for non-contiguous derivation is not working, but the results should be true ...)
\end{proof}

A \emph{permutation} on $\interval{n}$ is a bijection
$\perm : \interval{n} \to \interval{n}$. We call it a
\emph{transposition} when there are $i, j \in \interval{n}$,
$i \neq j$ such that $\sigma(i)=j$, $\sigma(j) = i$ and
$\sigma(k) = k$ otherwise. Such a transposition is normally denoted as
$\transp{i}[j]$. Whenever $j = i+1$ we call it an \emph{adjacent
  transposition}.

Given a permutation $\perm$ on $\interval{n}$, the set of inversions
of $\perm$ is
$\inv{\perm} = \{ (i,j) \mid i,j \in \interval{n-1}\ \land\ i<j\
\land\ \perm(i) > \perm(j) \}$. Clearly, $\perm$ is the identity iff
$\inv{\perm}=0$.

Now, if $\perm \neq id$ there is at least an inversion of the kind
$(i,i+1)$.  Then we have that $\perm = \transp{i}; \perm'$
with $|\inv{\perm'}| = |\inv{\perm}| -1$. Hence we can inductively
express $\perm$ as a (minimal length) composition of adjacent
transpositions.

\begin{lemma}[on permutations]
  \label{le:perm}
  Let $\perm$, $\perm'$ be permutations on $\interval{n}$ and
  $\perm = \upsilon_1; \perm'; \upsilon_2$ where $\upsilon_1$ and $\upsilon_2$
  are transpositions with $\upsilon_1 = \transp{i}[j]$ and
  $\upsilon_2 = \transp{\perm'(i)}[\perm'(j)]$. Then $\sigma =
  \sigma'$.

  Moreover, let $\upsilon_1, \upsilon_2$ be adjacent
  transpositions and $\sigma' = \upsilon_1; \ldots, \upsilon_m$ a
  minimal length decomposition of $\sigma'$ in terms of adjacent
  transpositions (that thus operate on inversions). Then, for all
  $h \in \interval{m}$ with $\upsilon_h = \transp{l}[l+1]$ there exists
  $k \in \interval{m}$ such that
  $\upsilon_k = \transp{\upsilon_{1,h-1}^{-1};v_1;
    \upsilon_{1,k-1}(l)}[\upsilon_{1,h-1}^{-1};v_1; \upsilon_{1,k-1}(l+1)]$.
\end{lemma}

\begin{proof}
  The first part is easy. The second is a consequence of the fact that
  the minimal decomposition in terms of adjacent transpositions is
  essentially unique, in the sense that it consists of exchanges of
  all the inversions. To be formalised.
\end{proof}

In words, the lemma above says that if a permutation is decomposed as $\sigma = \upsilon_1; \sigma'; \upsilon_2$ with $\sigma' = \upsilon_1'; \ldots, \upsilon_n'$, where $\upsilon_1$ and $\upsilon_2$ performs a switch and its reverse, and the inner decomposition is minimal, then we can strip the first and last switch and get a decomposition of the same permutation that performs the same switches but possibly in a different order.

\begin{lemma}[no need of useless shifts]
  \label{le:no-useless-shifts}
  Let $\rho$, $\rho'$ be direct derivations and assume
  $\rho \shiftpre[\sigma] \rho'$. Then
  $\rho = \rho_0 \shiftdir[\upsilon_1] \rho_1 \ldots \shiftdir[\upsilon_n ]
  \rho'$ where $\sigma = \upsilon_1; \ldots; \upsilon_n$ and $\upsilon_j \in \inv{\upsilon_{j,n}}$ for $j \in \{1, \ldots,n\}$.
\end{lemma}

\begin{proof}
  Let $\rho \shiftpre[\perm] \rho'$. This means that
  $\rho = \rho_0 \shiftdir[\upsilon_1] \rho_1 \ldots \shiftdir[\upsilon_n ]
  \rho'$ with $\perm = \upsilon_1; \ldots; \upsilon_n$ for some $n \geq 0$.

  We show that if some $\upsilon_j$ does not act on an inversion, i.e.,
  $\upsilon_j \not\in \inv{\sigma_{j,n}}$ then we can shorten the
  proof of $\rho \shiftpre[\sigma] \rho'$ of two steps. This will
  allow us to conclude that we can reduce it to a proof only acting
  on inversions, as desired.

  The key observation is that, if
  $\upsilon_j \not\in \inv{\sigma_{j,n}}$ then, intuitively, the ``same
  exchange'' must be performed in the reverse direction by some
  $\upsilon_h$ ($h>j$).  Formally, if $\upsilon_j = \transp{i}$ and
  $(i,i+1)\not\in\inv{\upsilon_{i,n}}$ then there exists $h>j$ such that
  $\upsilon_{j,h-1}(i) = \upsilon_{j,h-1}(i+1)+1$ and
  $\upsilon_h = \transp{\upsilon_{j,h-1}(i+1)}$. Among all the pairs
  $j, h$ of this kind, take one with minimal $h-i$. This means that in
  the middle all transpositions are inversions of $\upsilon_{i+1,h-1}$.

  By Lemma~\ref{le:perm}, $\upsilon_i$ and $\upsilon_h$ can be
  stripped, i.e., $\upsilon_{i,h} = \upsilon_{i+1,h-1}$. Moreover,
  again by Lemma~\ref{le:perm},
  $\upsilon_{i+1}, \ldots, \upsilon_{h-1}$ perform exchanges that were
  already done in the original proof, and thus, since independence is
  invariant under shift equivalence, as expressed by
  Lemma~\ref{le:seq-ind-glob1} they are allowed. Moreover, since by
  Lemma~\ref{le:unique-IC} the result of switching is uniquely
  determined (up to iso), the resulting derivation is the same as
  before, hence we can strip $\upsilon_{i}$ and $\upsilon_{h}$,
  and conclude.
\end{proof}

Then we can prove that whenever $\rho \shiftpre[\perm] \rho'$, each
decomposition of $\perm$ in terms of inversions leads to a valid
derivation of $\rho'$ from $\rho$.

\begin{lemma}[independence is global/2]
  \label{le:seq-ind-glob2}
  Let $\rho$, $\rho'$ be direct derivations. Let
  $\rho \shiftpre[\perm] \rho'$, with $\perm \neq id$ and let
  $\perm = \transp{i}; \perm'$ where $(i,i+1) \in \inv{\perm}$. Then
  there exists $\rho''$ such that
  $\rho \shiftpre[\transp{i}] \rho'' \shiftpre[\perm']
  \rho'$.
\end{lemma}

\begin{proof}
  Let
  $\rho = \rho_0 \shiftdir[\upsilon_1] \rho_1 \ldots
  \shiftdir[\upsilon_n] \rho'$ where $\sigma = \upsilon_1; \ldots; \upsilon_n$ and $\upsilon_j \in \inv{\upsilon_{j,n}}$ for $j \in \{1, \ldots,n\}$, be the decomposition
  provided by Lemma~\ref{le:no-useless-shifts}.

  We know that steps corresponding to $i$ and $i+1$ are exchanged at
  some point, namely there is $j$ such that
  $\upsilon_j =
  \transp{\upsilon_{1,j-1}(i)}[\upsilon_{1,j-1}(i+1)]$. By
  Lemma~\ref{le:seq-ind-glob1} the steps $i$ and $i+1$ are independent
  already at the beginning and thus we can exchange them first.
  More precisely, 
  \begin{center}
    $\rho \shiftdir_{\transp{i}} \rho'' \shiftdir_{\upsilon_{1,j-1}}
    \rho_j \shiftdir_{\upsilon_{j+1,n}} \rho'$.
  \end{center}
  This can be proved by induction on $j$.
\end{proof}

\begin{lemma}[more on shift]
  \label{le:more-shift}
  if $\rho_1;\rho_1' \shiftpre[\perm | \perm'] \rho_2; \rho_2'$,
  with $|\rho_1| = |\rho_2|$ and
  $\sigma : \interval{|\rho_1|-1]} \to \interval{|\rho_2|-1]}$ then
  $\rho_1 \shiftpre[\perm] \rho_2$ and
  $\rho_1' \shiftpre[\perm'] \rho_2'$.
\end{lemma}

\begin{proof}
  The results follows by the fact that, from
  Lemma~\ref{le:no-useless-shifts}, we derive that
  \begin{center} 
    $\rho_1;\rho_1' \shiftdir[ \upsilon_1 \mid id_{[1..|\rho_1'|]}]
    \ldots \shiftdir[ \upsilon_h \mid id_{[1..|\rho_1'|]}] \rho_2;
    \rho_1' \shiftdir[ id_{[1..|\rho_1|]} \mid \upsilon_1'] \ldots
    \shiftdir[ id_{[1..|\rho_1|]} \mid \upsilon_k'] \rho_2; \rho_2'$
  \end{center}
  where $\upsilon_1; \ldots; \upsilon_h$ is a decomposition of $\perm$
  in terms of inversions and, similarly,
  $\upsilon_1'; \ldots; \upsilon_k'$ is a decomposition of
  $\perm'$. Then
  $\rho_1 \shiftdir[\upsilon_1] \ldots \shiftdir[\upsilon_h] \rho_2$
  and analogously
  $\rho_1' \shiftdir[\upsilon_1'] \ldots \shiftdir[\upsilon_k']
  \rho_2'$, which leads to the desired conclusion.
\end{proof}

\todo[inline]{Note that above we assume that $\rho_1$ and $\rho_1'$ ends in the same graph. To be formalised.}

\begin{lemma}[uniqueness of permutation]
  if $\rho \shiftpre[\perm_1] \rho'$ and
  $\rho \shiftpre[\perm_2] \rho'$ then $\perm_1 = \perm_2$. In
  particular, if $\rho_1 \shiftpre[\perm_1] \rho_1$ then $\perm_1 = id$.
\end{lemma}

\begin{proof}
  We proceed by induction on the length of the derivations
  $\ell=|\rho|=|\rho'|$ and by an inner induction on the number of
  inversions in $\sigma_1$. When $\ell = 0$, the result is
  trivial. Let us assume $\ell>0$ and let $i_1 = \sigma_1^{-1}(1)$,
  $i_2 = \sigma_2^{-1}(1)$. Let
  $L_1' \stackrel{l_1'}{\leftarrow} K_1' \stackrel{r_1'}{\rightarrow}
  R_1')$ be the rule used by the first production in $\rho'$.
  By Lemma 17(1) in the paper [the implication
  that holds] $\sigma_1$ and $\sigma_2$ are consistent permutations of
  $\rho$ into $\rho'$.
  %
  Since they are consuming, it is easy to see that
  $i_1=\sigma_1^{-1}(1)$ is uniquely determined by the (equivalence
  class) of an item in the initial graph consumed by $L_1'$ and the
  same applies to $i_2=\sigma_2^{-1}(1)$, hence we conclude that
  $i_1=i_2$.

  Now, if $i_1=i_2=1$, then $\sigma_j = \transp{1}[1] \mid \sigma_j'$
  for $j \in \{1,2\}$ and $\rho = \delta; \rho_1$,
  $\rho' = \delta; \rho_1'$ where $\delta$ is the first direct
  derivation, and $\rho_1 \shiftpre[\perm_j'] \rho_1'$ for
  $j \in \{1,2\}$. Since $|\rho_1| = \ell-1$ we conclude by inductive
  hypothesis that $\sigma_1'=\sigma_2'$ and thus $\sigma_1=\sigma_2$,
  as desired.

  If instead $i_1=i_2=i>1$, since $\transp{i_1-1}[i_1]$ is an inversion for both
   $\sigma_1$ and $\sigma_2$, by Lemma~\ref{le:seq-ind-glob2} we can
  perform such a shift first, obtaining
  \begin{center}
    $\rho \shiftdir[{\transp{i-1}[i]}] \rho_1
    \shiftpre[{\transp{i-1}[i];\sigma_1}] \rho'$
    %
    \quad and \quad
    %
    $\rho \shiftdir[{\transp{i-1}[i]}] \rho_1
    \shiftpre[{\transp{i-1}[i];\sigma_2}] \rho'$
  \end{center}
  Since ${\transp{i-1}[i];\sigma_1}$ has an inversion less than
  $\sigma_1$ we conclude by inductive hypothesis that
  ${\transp{i-1}[i];\sigma_1} ={\transp{i-1}[i];\sigma_2}$, whence
  $\sigma_1=\sigma_2$.
\end{proof}

Note that the above works also with the liberal definition of
sequential independence.

\begin{lemma}[Lemma 17(2) in the paper]
  if $\rho;\rho_1 \shiftpre[\perm] \rho; \rho_2$ then
  $\perm = id_{[1..|\rho|]} \mid \perm'$ and
  $\rho_1 \shiftpre[\perm'] \rho_2$.
\end{lemma}

\begin{proof}[skecth]
  Productions are `uniquely identified by the items they consume,
  hence productions in $\rho$ are necessarily mapped to themselves,
  i.e., $\perm = id_{[1..|\rho|]} \mid \perm'$. Then we conclude
  $\rho_1 \shiftpre[\perm'] \rho_2$ by Lemma~\ref{le:more-shift}.
\end{proof}

\begin{lemma}[Lemma 20 in the paper]
  \label{le:prop-c}
  Let $\psi$ and $\psi'$ be decorated derivations. Then the following hold:
  \begin{enumerate}
  \item
    \label{le:prop-c:1}
    Let $\psi_1, \psi_1'$ be such
    that $\psi;\psi_1 \equiv^c_{\perm} \psi';\psi_1'$ and let
    $n = | \{ j \in \interval[|\psi|]{|\psi;\psi_1|-1} \mid \perm(j)
    < |\psi'| \}|$. Then for all $\phi_2, \phi_2'$ such that
    $\psi;\phi_2 \equiv^c \psi';\phi_2'$ it holds $|\phi_2| \geq n$
    and there are $\psi_2, \psi_2',\psi_3$ such that

    \begin{itemize}
    \item $\psi;\psi_1 \equiv^c \psi;\psi_2;\psi_3$
    \item $\psi;\psi_2 \equiv^c \psi';\psi_2'$

    \item $|\psi_2| = n$

  \item
    \label{le:prop-c:2}
    Let $\psi_1, \psi_1'$, $\psi_2, \psi_2'$ be such that
    $\psi;\psi_1 \equiv^c_{\perm_1} \psi';\psi_1'$ and
    $\psi;\psi_2 \equiv^c_{\perm_2} \psi';\psi_2'$ with $\psi_1, \psi_2$
    of minimal length. Then $\psi_1 \equiv^c_\perm \psi_2 \cdot \nu$,
    where $\nu : \target{\psi_2} \to \target{\psi_2}$ is some graph isomorphism and
    $\perm(j) = \perm_2^{-1}(\perm_1(j + |\psi|))-|\psi|$ for
    $j \in \interval[0]{|\psi_1|-1}$.    
  \end{enumerate}
\end{lemma}

\begin{proof}[sketch]  
  Point ~\ref{le:prop-c:1}: The proof should be similar to the one we have already, basing the rearrangement on Lemma~\ref{le:seq-ind-glob2}. Roughly we get $\psi;\psi_1 \equiv^c \psi_a;\psi_b;\psi_{2};\psi_{3}$ and
  $\psi';\psi_1' \equiv^c \psi_a';\psi_b';\psi_{2}';\psi_{3}'$
  with
  \begin{center}
    $\psi_a;\psi_b;\psi_{2};\psi_{3}
    \equiv^c_{\perm_a|\perm_{b2}|\perm_2}
    \psi_a';\psi_b';\psi_{2}';\psi_{3}'$
  \end{center}
  with $\perm_a$ over $\interval{|\psi_a|-1}$, $\perm_{b2}$ over
  $\interval{|\psi_b;\psi_{2}|-1}$ ``mapping $\psi_b$ to $\psi_{2}'$
  and $\psi_b'$ to $\psi_{2}$'', $\perm_3$ over
  $\interval{|\psi_2|-1}$.  We deduce by Lemma~\ref{le:more-shift}
  that $\psi_a \equiv^c_{\perm_a} \psi_a'$,
  $\psi_b;\psi_{2} \equiv^c_{\perm_{b2}} \psi_b';\psi_{2}'$ and
  $\psi_{3} \equiv^c_{\perm_3} \psi_{3}'$.

  Now, for all $\phi_2, \phi_2'$ such that
  $\psi;\phi_2 \equiv^c \psi';\phi_2'$ we can perform the same
  splitting and this leads to the conclusion that $|\phi| \geq n$ and
  also it can be expressed as $\psi_2;\phi_2'$, that should subsume
  also the second point.
  %
\qed
\end{proof}

% \paragraph{Note.} This does not solve the problem raised by Andrea's counterexample that still applies.

Another idea could be to try to use the inductive characterisation in Figure~\ref{fi:ind} (but I do not see how to mange the transitive rule).

\begin{figure}
  \begin{center}
    \small
  \[
    \inferL{\slab{(SH-id)}}
    {\rho \shiftpre[id] \rho}
    {}
    % 
    \hspace{0.9cm}
    % 
    \inferL{\slab{(SH-exch)}}
    {\rho \shiftpre[{\transp{1}[2]}] \rho'}
    {\rho = \upsilon_1;\upsilon_2 & \rho' \in \IC{\rho}}
  \]

  \[
    \inferL{\slab{(SH-comp)}}
    {\rho_1 ; \rho_2  \shiftpre[\upsilon_1 \mid \upsilon_2] \rho'_1 ;
      \rho'_2}
    {\rho_1 \shiftpre[\upsilon_1] \rho'_1, \ \rho_2 \shiftpre[\upsilon_2] 
      \rho'_2,\ \tau(\rho_1) = \perm(\rho_2)}
  \]

  \[
    % 
    % \inferL{\slab{(SH-sym)}}
    % {\rho' \shiftpre[\perm^{-1}] \rho}
    % {\rho \shiftpre[\perm] \rho'}
    % % 
    % \hspace{0.3cm}
    % 
    \inferL{\slab{(SH-tr)}}
    {\rho \shiftpre[\perm'\circ \perm] \rho''}
    {\rho \shiftpre[\perm] \rho',\ \rho' \shiftpre[\perm'], \rho''}
    %
    \hspace{0.3cm}
    \inferL{\slab{(SH-eq)}}
    {\rho' \shifteq[\perm] \rho}
    {\rho \shiftpre[\perm] \rho' & \rho' \shiftpre[\perm^{-1}] \rho}
  \]

%  \noindent
%  for $\perm_1 \mid \perm_2$ the pairing of two permutations,
%  and similarly $\sqsubseteq^c_{\perm}$ is defined inductively by

  \[
    \inferL{\slab{(C-abs)}}
    {\psi \equiv^c_{id} \psi'} 
    {\psi \equiv^{a} \psi'} 
    % 
    \hspace{0.3cm}
    % 
    \inferL{\slab{(C-sh)}}
    {\langle \alpha, \rho, \omega \rangle \equiv^c_{\perm} \langle \alpha, \rho', \omega \rangle} 
    {\rho \shifteq[\perm] \rho'}
    % 
    \hspace{0.3cm}
    % 
    \inferL{\slab{(C-tr)}}
    {\psi \equiv^c_{\perm' \circ \perm} \psi''}
    {
      {\psi \equiv^c_{\perm} \psi'} & 
      {\psi' \equiv^c_{\perm'} \psi''}
    }
  \]
\end{center}
\caption{Inductive definition of $\shiftpre$ and $\equiv^c$, for $\perm_1 \mid \perm_2$ the pairing of two permutations.}
\label{fi:ind}
\end{figure}

>>>>

In particular, if one just take the definition with general
interchange pairs, undesidered things might happen. In particular one
can have shift-equivalent derivations where the permutation is the
identity but which are non isomorphic (e.g., in GraphShift, page 1-2
$\rho_1 \shifteq \rho_3$, but $\rho_1 \not\equiv^{a} \rho_3$).

We want a notion of shift-equivalence that ensures some properties
which are indispensable to allow to interpret derivations up-to shift
as equivalent concurrent computations which differ only for the fact
that independent steps have switched. When moving around steps, we
should not lose track of the identity of the steps: this is essential
to be able to have an event structure semantics with a clearly defined
notion of event.

Hence we assume to deal with a notion of shift-equivalence between
derivations of the kind $\rho \shifteq_\sigma \rho'$ where
$\rho, \rho'$ consist of $n$ rewriting steps,
$\sigma: \interval{n} \to \interval{n}$ is a permutation (the idea
being that $\rho$ is obtained from $\rho'$ permutating the steps
according to $\sigma$) with - at least - the following properties:

\begin{itemize}
\item \textbf{Property 0}. Shift-equivalence is an equivalence! In
  particular it is symmetric. I.e., if $\rho \shifteq_\sigma \rho'$
  then $\rho' \shifteq_{\sigma^{-1}} \rho$;

@@

\textbf{Question.} Taking the standard shift with generic interchange
pair, in the case of non-linear rules a counterexample to Property 1
is the one mentioned above.  Do we have a counterexample to Property
2? Not clear to me.

@@

\item For left-linear rules, the notion of shift equivalence should
  require: uniqueness of the interchange pair and symmetry. Show that
  this satisfies the above

\item Apparently, the satisfaction of the properties above has not
  been explicitly observed, not even for the case of linear rules. If
  so, it is worth to remark it.

\item Does the theory above apply to other cases. E.g., if one wants a
  proper notion of causality (avoiding or-causality) a possibility
  seems to be to restrict the applicability condition for rules, to
  prevent fusions to act on already merged items (see section V.D in
  UnstableFull). Incidentally, note that the satisfaction of the
  conditions on shift depends not only on the notion of shift, but
  also on the notion of rewriting.

  In this specific case we expect to have a prime event structure (but
  this has to do with the second part of the work).

@@

Ideally, one would like to be able to work with an abstract notion of
shift-equivalence and just using its properties as singled out in the
previous part to derive a weak prime domain or connected event structure
for it. This is probably difficult, but one might need to add some
further properties to be proved for the specific notion of shift at
hand.

For this part, relevant material is section V of UnstableFull. We use abstraction equivalence (iso of derivations) $\equiv^{a}$ (Def. 36), shift-equivalence $\shifteq_\sigma$ (Def. 38, tracking the permutation) and their join $\equiv^c_\sigma$.

For linear rules on graphs it is known that two derivations are
related by $\rho \equiv^c_\sigma \rho'$ iff $\sigma$ is a consistent
permutation (Def. 40, UnstableFull). Roughly the idea is to take the colimit
and consider the rewriting steps over the colimit to deduce
dependencies. This was used in many places in the proofs, but for
left-linear rules the proof does not go through (only the implication
``if $\rho \equiv^c_\sigma \rho'$ then $\sigma$ is a consistent
permutation'' (Lemma 17(1)) seems to hold as the original proof do not
to rely on linearity, this should be double-checked, but the converse
is false).

In general, the hope could be to get rid of using consistent
permutations and use instead more abstract properties of
shift-equivalence as discussed in the first part. However, I think
that using the valid implication (``if $\rho \equiv^c_\sigma \rho'$
then $\sigma$ is a consistent permutation'') can be quite useful to track the identity of rewriting steps, as they can be identified with the items consumed in the colimit.

The general idea for getting a domain is to show that considering derivations are arrows from the source to the target graph, then the slice from the initial graph $\slice{[G_s]}{\tr{\mathcal{G}}}$ should be a pre-order, whose ideal completion is a weak prime domain.

@@
